$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-04-29 paper/paper.tex
Give varref judgments a pass

\todo{rename Scoped to Process in code}
\todo{make sure all citations are there}
\todo{change colors, color functions}
\todo{do not distinguish fields from functions, it is not worth it}

@@

We choose to allow the use of multiple coexisting multiplicity algebras in our typing system.
We do so by defining an indexed set of usage algebras.

@@

\subsection{Typing with Leftovers}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\label{leftover-typing}

%%%% REARRANGE
In this paper, we combine typing and annotation judgments into a single judgment.
Although the typing rules are for now completely syntax-directed, adding support for branching would introduce choice in the typing derivations.
If we were to keep typing judgments and annotation judgments separate, we would need a coherence relation making sure that the same choice is taken by both.
We avoid doing so by combining them into a single judgment.
%%%%

We make use \emph{leftover typing}, a technique consisting of adding an \emph{output usage context} to the usual \emph{input usage context}.
This output context contains the leftovers: the multiplicities unused by the processes.
Crucially, this output context can then be used as the input context of the continuation process.
The typing rules are introduced in \S \ref{leftover-typing}.

We use leftover typing \cite{} to encode our typing rules.
The insight of this approach is to use input and output contexts in the typing judgment.
Our typing judgment on processes are of the form $\types{\gamma}{\Gamma}{P}{\Delta}$ (introduced in Figure \ref{fig:types}).
This assertion is in fact both a typing judgment on $\gamma$ and a usage judgment on $\Gamma$ and $\Delta$.
These two judgments could, in fact, happen separately: our typing rules are completely syntax directed.
However, if we were to introduce branching, typing derivations would have to make choices, and the two separate judgments would need of a third relation to make sure that their choices agree.
To avoid this, we merge both judgments into a single one.
The typing context $\gamma$ needs to be explicit because the usage context $\Gamma$ and $\Delta$ do not in any way depend on it -- in fact, they are only related by their common length.

While the typing contexts are preserved through the typing derivation, the usage contexts change.
We use input and output usage contexts to capture this transformation.
This has two main advantages: for one, the usage modifications effected by both input and output can be enclosed in a self-containing typing and usage judgment for variable references (also with an output context); but more crucially, we avoid requesting an explicit constructive proof of context splitting for parallel composition -- if $P$ is typed under $\Gamma_l$ and $Q$ under $\Gamma_r$, then to type $\comp{P}{Q}$ under $\Gamma_m$ a proof $\opctx{\Gamma_m}{\Gamma_l}{\Gamma_r}$ would be needed.
The search for such a proof can be automated, but its existence is still required.
Moreover, providing a typing derivation for $P$ gets the user no usage annotations to type $Q$.
By using output usage contexts, we get the leftovers from the derivation on $P$ and use them to type $Q$.
This is the recurring theme in leftover typing: usage contexts are threaded through the premises.

The typing and usage judgments on both processes and variable references are of an inductive form.
As a result, the proofs on typing derivations will perform a double induction: once on the derivation of the process, once on the derivation of the variable reference.

\begin{nidefinition}
  Akin to the typing and usage judgment on processes, in Figure \ref{fig:variable-references} we present a typing and usage judgment on variable references -- also indexed over a typing context and an input an output usage context.
  However, in addition to this, variable references also record the index, the type, and the usage annotation of the variable reference.
  The base case $\constr{zero}$ ensures that $t$ is the most immediate type in typing context, and splits the most immediate annotation in the input context $x$ into the usage annotation for the variable $y$ and the annotation in the output context $z$.
  The inductive case $\constr{suc}$ appends an extra type to the typing context, and an extra usage annotation to both the input and output usage context.

@@

    \inferrule
    {loc_i : \contains{\gamma}{\Gamma}{i}{t}{x}{\Delta}}
    {\constr{suc} \; loc_i : \contains{\gamma \comma t'}{\Gamma \comma x'}{\suc i}{t}{x}{\Delta \comma x'}}
  \end{mathpar}
  \caption{Variable references as proofs on the typing context and input and output multiplicity contexts.}
  \label{fig:variable-references}
  \end{figure}
\end{nidefinition}

\begin{remark}
  We use $\field{\cdot-unique}$ to relieve the user from the proof burden $\opsquared{x}{y}{z}$ in the base case $\constr{zero}$.
  This is the only place in the typing system that uses the algebra on multiplicities.
\end{remark}

@@

\begin{nilemma}
  The typing context $\gamma$ at index $i$ is type $t$.
\end{nilemma}
\begin{nilemma}
  The input annotation in context $\Gamma$ at index $i$ is the result of aggregating (as per the algebra) $y$ and the output annotation in context $\Delta$ at index $i$.
\end{nilemma}
\begin{nilemma}
  For every index $j$ other than $i$, the input annotation in context $\Gamma$ at index $i$ is the output annotation in $\Delta$ at index $i$.
\end{nilemma}
\begin{nilemma}
  \todo{Add feedback and feedfront}
\end{nilemma}
\begin{corollary}
  Given the witness $\contains{\gamma}{\Gamma}{i}{t}{y}{\Delta}$, $\Gamma \equiv \Delta$ iff $y \equiv \zero$.
\end{corollary}
\proof{All proofs are by structural induction on \textsc{VarRef}.}

\begin{nidefinition}
  The typing relation \textsc{Types} presented in Figure \ref{fig:types} is indexed over a context of types $\gamma$, a context of input multiplicities $\Gamma$, the process $P$ that is being typed, and the context of output capabilities $\Delta$, and notated as $\types{\gamma}{\Gamma}{P}{\Delta}$.
  \todo{empty output context}
  The typing rules are in one-to-one correspondence with the underlying syntax for processes, and thus we say they are syntax-directed.

>>>>

\todo{rename Scoped to Process in code}
\todo{make sure all citations are there}
\todo{change colors, color functions}
\todo{do not distinguish fields from functions, it is not worth it}
\todo{go over guillaume's paper again}

@@

We choose to allow multiple multiplicity algebras to coexist in our typing system.
We do so by defining an indexed set of usage algebras.

@@

\subsection{Typing with Leftovers}
\label{leftover-typing}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We make use \emph{leftover typing} \cite{}, an approach that, in addition to the usual typing context and input usage context, adds an extra output usage context to the typing rules.
This output context contains the \emph{leftovers} of the process being typed --- its unused multiplicities.
Crucially, the leftovers can then be used as an input to some othew typing derivation.
Adopting this techinque has two main advantages:
\begin{itemize}
  \item
    The input and output actions do no longer need to transform the usage context through functions, but can use a variable reference type with both an input and an output usage context to encapsulate these transformations.
  \item
    Parallel composition no longer needs an explicit context splitting proof --- a proof $\opctx{\Gamma_m}{\Gamma_l}{\Gamma_r}$ where $\Gamma_l \vdash P$, $\Gamma_r \vdash Q$ and $\Gamma_m \vdash \comp{P}{Q}$.
    Such a proof could be obtained through automated search --- but it is better yet to avoid needing it.
    With leftover typing, we use the leftovers of $P$ to type $Q$.
\end{itemize}

We introduce two judgments: one for variable references (Definition \ref{def:varref}), one for processes (Definition \ref{def:types}).

\begin{nidefinition}
  \label{def:varref}

  Like processes, variable reference judgments (Figure \ref{fig:variable-references}) are indexed by a typing context and an input and output usage context.
  In addition to these, variable reference judgments are also indexed by their position in the contexts, their type, and their usage annotation.

  The base case $\constr{zero}$ asserts that $t$ is the most immediate type in typing context, and splits the most immediate usage annotation $x$ in the input context into the usage annotation $y$ the usage annotation $z$ in the output context.
  This is the only place in the typing system where the algebra on multiplicities is used.

  \begin{remark}
    We use $\field{\cdot-compute}$ to aliviate the user from the proof burden $\opsquared{x}{y}{z}$.
  \end{remark}

  The inductive case $\constr{suc}$ appends an extra type to the typing context, and an extra usage annotation to both the input and output usage context.

@@

    \inferrule
    {loc_i : \contains{\gamma}{\Gamma}{i}{t}{x}{\Delta}}
    {\constr{suc} \; loc_i : \contains{\gamma \comma t'}{\Gamma \comma x'}{\suc i}{t}{x}{\Delta \comma x'}}
  \end{mathpar}
  \caption{Variable reference judgments.}
  \label{fig:variable-references}
  \end{figure}
\end{nidefinition}

@@

\begin{nidefinition}
  \label{def:types}

  Our judgment on processes is of the form $\types{\gamma}{\Gamma}{P}{\Delta}$ (Figure \ref{fig:types}).
  This assertion combines a judgment on types ($\gamma$) and a judgment on usages ($\Gamma$ and $\Delta$).
  They could be kept as two separate judgments, but doing so would risk incoherence: if there is any choice to be made, the judgment on types could choose to go left while the judgment on usages chooses to go right.
  The constructions instroduced in this paper make the typing derivations completely syntax directed and thus this problem inexistent, but we nevertheless choose to keep both judgments as one for simplicity.

  The typing relation \textsc{Types} presented in Figure \ref{fig:types} is indexed over a context of types $\gamma$, a context of input multiplicities $\Gamma$, the process $P$ that is being typed, and the context of output capabilities $\Delta$, and notated as $\types{\gamma}{\Gamma}{P}{\Delta}$.
  \todo{empty output context}
  The typing rules are in one-to-one correspondence with the underlying syntax for processes, and thus we say they are syntax-directed.
